Calibrating MEM seeding heuristics

### 2019-01-26 MEM.tex
Added references, started to write a clear document.

\subsection{Mapping}

Consider the following task referred to as the \emph{true mapping
problem}: a random DNA fragment from a known genome is sequenced with an
imperfect instrument; can we identify the location of the fragment in the
genome?

It seems that the answer depends only on the size of the fragment and the
error rate of the instrument, but there is another important factor to
consider. Most genomes contain repeated sequences. Intuitively, the
problem is more difficult if the random DNA fragment is duplicated. If one
of the duplicates is exactly identical to the original, then the DNA
fragment cannot be mapped with certainty. Even if the duplicates are all
different from the original (which happens because of mutations in the
genome), the read may still be closer to one of the duplicates because of
sequencing errors. So we can never be sure that the read is mapped to the
\emph{true} location.

For this reason, we introduce the \emph{best mapping problem}, where we
search not the true, but the \emph{best} location of the DNA fragment,
\textit{i.e.} the sequence of the genome that is the most similar to the
read.

There are exact solutions to the best mapping problem, but the algorithms
are too slow to process the large amount of data generated by modern
sequencers. Instead, one uses heuristic methods, \textit{i.e.} algorithms
that run fast, but are not guaranteed to return the correct answer. In the
context of the best mapping problem, there are thus three possible
outcomes: $i.$ the DNA fragment is mapped to the best location, which is
called a \emph{true positive} (or an \emph{on-target hit}), $ii.$ the DNA
fragment is mapped to another location, which is called a \emph{false
positive} (or an \emph{off-target hit}), $iii.$ the DNA fragment is not
mapped at all, which is called a \emph{false negative}. A \emph{true
negative} would designate the outcome where a DNA fragment foreign to the
genome is not mapped at all, but here we assume that all DNA fragments
belong to the genome of interest.

In a seeding heuristic, one uses exact or near exact matches between the
read and the genome to collect a set of candidate locations. The read is
only compared to the sequences at the candidate locations, but not to the
rest of the genome. The advantage of this method is that there are few
candidates and that they can be discovered fast. However, the best hit may
not be in the candidate set, in which case it cannot be discovered
(\textit{i.e.}, the outcome is either a false positive or a false
negative).

A \emph{seed} designates a subsequence of fixed length that is shared
between the read and the genome. In other words, a seed is a perfect match
between a part of the read an a genomic site. The candidates consist of
the genomic sites that contain one or more seeds (this depends on the
specific heuristic).

In the theory developed below, we will assume that the \emph{true}
location of the read is also the \emph{best}. This is not always the case,
but it is unfeasible to model the behavior of the best hit in the present
context. We will later discuss how this relates to the probability that
the best hit has a seed.

\subsection{Maximal Exact Matches (MEMs)}

Here we consider a seeding variant where seeds do not have a fixed
length. Let us start by introducing the definition of a Maximal Exact
Match.

\begin{definition}
A Maximal Exact Match (MEM) is a subsequence of a read that is present in
the genome and that cannot be extended left or right --- either because
the read ends or because the extended subsequence is not in the genome.
\end{definition}

This definition presents a computational challenge: to know if a given
subsequence is a MEM, we need to know if it exists somewhere in the
genome. This is a non trivial problem in itself, but fortunately there
exists practical methods to answer the question, even for very large
genomes. Here on, we will assume that we always know which subsequences of
the read are MEMs.

A practical consideration is that short matches between the read and the
genome are irrelevant. For instance, $99.7\%$ of all the possible 12-mers
are present in the reference human genome, so an exact match of size 12 or
lower says nothing about the location of the DNA fragment. In what
follows, we will simply disregard short matches. More specifically, we
will assume that subsequences of the read can only match the original DNA
sequence or its duplicates, but not the rest of the genome. This will
simplify the discussion without loss of generality, because short
uninformative matches are ignored in all practical applications.

Fig.~\ref{fig:MEM_example} shows a concrete example of the relationship
between read errors, on-target matches, off-target matches and MEMs. The
sequence of the original DNA fragment is the first genomic hit (this is
the \emph{true} location). The second, third and fourth genomic hits are
duplicates. The mismatches between the genome and the read are
highlighted in bold. The read has three errors indicated by stars, so
\texttt{T}, \texttt{T} and \texttt{C} should actually be \texttt{A},
\texttt{A} and \texttt{G} (because the first hit is the \emph{true}
sequence).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{MEM_example.pdf}
\caption{\textbf{Errors, matches and MEMs.}
Some legend.}
\label{fig:MEM_example}
\end{figure}

The MEMs are indicated by grey boxes. Technically, MEMs are subsequences
of the read, but they are shown on the genomic hits to highlight where
they match. Two MEMs can overlap, in which case they must match distinct
sequences of the genome (this follows from the definition). The last MEM
corresponds to two matches of length 4, with the first and the second
genomic hits. Such MEMs with several matches are referred to as
\emph{shared}. MEMs with a single match in the genome are referred to as
\emph{strict}.

As the name suggests, the principle of MEM seeding is to use MEM as seeds
to identify candidate genomic locations. Because short matches are
irrelevant, MEM seeds are usually requested to have a length greater than
or above a certain threshold denoted $\gamma$. We will assume that a
single MEM seed is sufficient to consider that a genomic hit is a
candidate location, which is why we will focus on the case that the read
does not have a single on-target MEM seed.

%In the example of Fig.~\ref{fig:MEM_example}, there is no MEM that matches
%the third genomic hit. It is important to understand that this can also
%happen to the best hit, as in the example shown in
%Fig~\ref{fig:MEM_eclipse}. The undesirable consequence is that some reads
%are bound to yield false positives, even if we tune the MEM seeding
%heuristic by changing the minimum seed length $\gamma$. We will refer to
%the case that the best hit is masked by off-target MEMs as \emph{MEM
%eclipse}. 
%
%\begin{figure}[h]
%\centering
%\includegraphics[scale=1]{MEM_eclipse.pdf}
%\caption{\textbf{MEM eclipse.}
%Some legend.}
%\label{fig:MEM_eclipse}
%\end{figure}
%
%If you are not already familiar with MEM seeding, we recommend that you
%take a moment to study the examples of Fig.~\ref{fig:MEM_example} and
%\ref{fig:MEM_eclipse}. This will be essential to understand the next
%sections. Before reading further, make sure that you can answer the
%following questions:
%\begin{enumerate}
%\itemsep0em
%\item What are the conditions for a MEM to be shared?
%\item What is the minimum number of duplicates for MEM eclipse?
%\end{enumerate}

@@

In this section we give a representation of reads as combinatorial ojects.
The construction has two essential purposes: the first is to give a
generative model of reads without any on-target MEM seed, the second is to
allow us to find their weighted generating function, which in turn will
allow us to find their probability of occurrence.

@@

The nucleotide sequence of the read is not so useful to know where MEMs
are and whether they are on-target. To develop a formal method to compute
the probability that the read contains an on-target MEM seed, we will
recode reads as sequences of letters from the so-called \emph{MEM
alphabet} $\{\downarrow, \uparrow, \square\}$.

Each nucleotide of the read is recoded with one of three letters. The
$\downarrow$ symbol indicates that the nucleotide is a sequencing error.
By definition, it is a mismatch for the target. The $\uparrow$ symbol
characterizes the presence of an on-target MEM in a way that we are about
to clarify. All the other nucleotides are represented by the $\square$
symbol.

To understand the meaning of the $\uparrow$ symbol, consider the following
situation: at the left end of the read, as we try to extend the first
match as much as possible to the right, the number of hits in the genome
keeps decreasing. It may be the case that at some point, we are left with
a single hit that is on-target. The very first nucleotid where that
happens is called the \emph{pivot} and is labelled with the symbol
$\uparrow$.

Fig.~\ref{fig:sketch_MEM} shows the read of Fig.~\ref{fig:MEM_example} as
encoded with the MEM alphabet. Below the read are the matches between the
read and the genomic hit, as they are in Fig.~\ref{fig:MEM_example}. A
black square means that the nucleotide is a mismatch for the genomic
sequence, and a white or grey square means that it is a match. Grey
squares are used to highlight MEMs, as in Fig.~\ref{fig:MEM_example}.

\begin{figure}[h]
\centering
\includegraphics[scale=.85]{sketch_MEM.pdf}
\caption{\textbf{Reads and MEM seeds}. 
Some legend.}
\label{fig:sketch_MEM}
\end{figure}

Observe that the two on-target MEMs contain a pivot, but the shared MEM
does not. This is always the case. Recall that the pivot indicates that
the true location is the \emph{only} local match, so the $\uparrow$ symbol
cannot appear in on-target MEM seeds when they are shared.

Strict on-target MEM seeds are easy to find when reads are represented in
the MEM alphabet: they are the longest stretches of symbols containing the
$\uparrow$ symbol and not containing the $\downarrow$ symbol. Shared
on-target MEM seeds are trickier. They do not directly appear in this
representation so we will have to keep that in mind throughout the theory.
It is in a sense justified to put more emphasis on strict MEM seeds
because they guarantee that the true location will be found. Shared MEM
seeds can have thousands of hits in the genome in practice, so the user
may not test them all. That said, the theory developed here fully takes
shared MEM seeds into account, even if they are not visible in the
representation of the MEM alphabet.

The point of the MEM seeding alphabet is that reads can be seen as
sequences of letters, but also as sequences of \emph{segments} delimited
by the $\uparrow$ and the $\downarrow$ symbols. Because such segments are
intimately linked to the presence of on-target MEM seeds, we can give a
construction of reads that do not contain any.

\begin{definition}
\label{def:segment}
A segment is a sequence of $\square$ symbols terminated by either a
$\uparrow$ symbol or by a $\downarrow$ symbol. The unterminated sequence
of $\square$ symbols at the end of the read is called the tail --- which
can thus be an empty sequence of symbols.
\end{definition}

Following definition~\ref{def:segment}, the $\uparrow$ and $\downarrow$
symbols will be referrred to as \emph{terminators}.
Fig.~\ref{fig:sketch_segment} shows the decomposition of the read from
Fig.~\ref{fig:sketch_MEM} in segments. The size of each segment (the
number of nucleotides that compose it) is indicated below.

\begin{figure}[h]
\centering
\includegraphics[scale=.85]{sketch_segments.pdf}
\caption{\textbf{Reads as sequences of segments}. 
Some legend.}
\label{fig:sketch_segment}
\end{figure}

@@

An issue with the MEM alphabet is that the probabilities of the symbols
are difficult to evaluate. The $\downarrow$ symbol coincides with read
errors, which is straightforward to model. The $\uparrow$ symbol coincides
with pivots, which have a more complex behavior. We thus extend the MEM
alphabet so that the occurrence of the $\uparrow$ becomes easier to
evaluate.

Only the terminators are extended; the $\square$ symbols remain unchanged.
Let us denote the number of duplicates of the target by $N$. The
$\downarrow$ symbol is replaced by the $N+1$ symbols
$\downarrow_{/m}$, where $m$ is the number of duplicates that
\emph{match} the erroneous nucleotide $(0 \leq m \leq N)$. The $\uparrow$
symbol is replaced by the symbols $\uparrow_{/j}$, where $i
\geq 0$ is the size of the segment (including the
$\uparrow_{/j}$ symbol). Finally, we add the special terminator
$|$ that marks the end of the read. In other words, tails are considered
normal segments terminated by the symbol $|$ (with size 0).

It is important to highlight that the numbers that decorate the symbols
have a very different nature. In the case of the $\downarrow$ symbol, it
is an information about the number of sequences matching the read; in the
case of the $\uparrow$ symbol, it is the position of the pivot relative to
the previous terminator (or to the beginning of the read).

Fig.~\ref{fig:sketch_extended} shows the decomposition of the read from
Fig.~\ref{fig:sketch_segment} in segments using the extended MEM alphabet.
The numbers decorating the symbols are shown above them to avoid crowding
the figure.

\begin{figure}[h]
\centering
\includegraphics[scale=.85]{sketch_extended.pdf}
\caption{\textbf{Reads as sequences of segments in the extended MEM
alphabet}. 
Some legend.}
\label{fig:sketch_extended}
\end{figure}

There is an infinite amount of $\uparrow_{/j}$ symbols in the
extended MEM alphabet, but if the read has no on-target MEM seed, then
only the $\uparrow_{/j}$ symbols where $1 \leq i \leq \gamma-1$
can be used (if a pivot terminates a segment of size greater than or equal
to $\gamma$, then the read contains a strict on-target MEM seed). The
extended MEM alphabet is made in such a way that the numbers decorating
the $\uparrow$ symbol will allow us to exclude strict on-target MEM seeds,
and the numbers decorating the $\downarrow$ symbol will help us exclude
shared on-target MEM seeds.

@@

@@

Let $\mathcal{A}$ be a set of combinatorial objects which have a
\emph{probability of occurrence}, and a \emph{size} that is an integer
number. The function defined as $A(z) = \sum_{k=0}^\infty a_kz^k$, where
$a_k$ is the total probability of objects of size $k$, is called the
\emph{weighted generating function} of the combinatorial objects.

In our case, we are interested in the weighted generating function of
reads without on-target MEM seed, where the coefficient $a_k$ is the
probability that a read of size $k$ has no on-target MEM seed. As $k$
becomes larger, it becomes more difficult to evaluate $a_k$, but the
combinatorial construction developed above will allow us to do this by
using established strategies involving weighted generating functions.

@@

The combinatorial construction of reads using the extended MEM seeding
alphabet gives us a method to express their weighted generating function
from that of segments. More specifically, we can describe complex
sequences of segments through a so-called \emph{transfer matrix} $M(z)$
that specifies which segments can follow one another in the reads of
interest.

The segments types are identified by their terminators and arranged in a
predefined oder. The entry of $M(z)$ at coordinates $(i,j)$ is the
weighted generating function of segments with the $j$-th terminator that
can be inserted immediately after segments with the $i$-th terminator. The
matrix $M(z) + M(z)^2 + \ldots = M(z) \cdot (I-M(z))^{-1}$ contains the
weighted generating functions of the reads of all sizes. The entry that
corresponds to terminators $\downarrow_{/0}$ and $|$ is the weighted
generating function of the reads of interest (the beginning of the read is
equivalent to a ghost segment terminated by $\downarrow_{/0}$,
\textit{i.e.} a mismatch for all the sequences).

With the transfer matrix $M(z)$, with thus have a way of computing the
weighted generating function of the reads without on-target MEM seed, and
from there we can extract teir probability of occurrence. Our next task is
thus to specify $M(z)$.

\section{The transfer matrix}

\subsection{Match streak and masks}

For each sequence, we define the \emph{match streak} at a given nucleotide
as the number of nucleotides since the last mismatch on the left. For
instance, in Fig.~\ref{fig:MEM_example}, the match streak of the first
genomic hit at the ninth nucleotide is $1$ while for the second genomic
hit, it is $7$.

>>>>

\subsection{Mapping reads to genomes}

Suppose that a DNA fragment drawn at random from a known
genome\footnote{We will assume throughout that the sequence of the genome
is known exactly and with absolute certainty. This is not the case in
practical application, but this assumption will allow us to develop a
theory centered on the read.} is sequenced with an imperfect instrument;
how can we identify the location of the fragment in the genome?

We will refer to this question as the \emph{true location problem}.
It is central to countless applications, including discovering
disease-causing mutations; diagnosing viral, bacterial or fungal
infections; diagnosing cancer and choosing appropriate therapies;
selecting breeds of interest in agriculture; tracing human migrations from
ancient bones; identifying the donor of a biological sample; diagnosing
genetic diseaes at birth; finding compatible graft donors \textit{etc}.

The focus of this document is solve the true location problem \textit{per
se}, but rather to evaluate whether a solution is correct. At first sight,
whether we can map reads to their true location seems to depend only on
their lengths and on the error rate of the instrument. However, there is
another important factor to consider: most genomes contain repeated
sequences, \textit{i.e.} relatively large subsequences (from $\sim20$~bp
to more than 10~kb) that are present at more than one location. So we
cannot map the fragment with certainty if there exists an exact copy of
the sequence at somewhere else in the genome, because we cannot know which
of the two copies was sequenced.

This case is rare though. Repeated sequences are in general not identical,
but merely homologous\footnote{By that we mean that drawing nucleotides
uniformly at random is very unlikely to produce sequences with the
observed degree of identity.}. So when the DNA fragment is duplicated, we
can map it with high confidence provided we can distinguish it from its
duplicates. This in turn depends on their similarity with the target
fragment and on the error rate of the sequencer.

Because sequencing errors can convert a DNA fragment into one of its
duplicates, the \emph{true} location of a DNA fragment is not always the
\emph{best} (as measured by the identity between the sequence and the
candidate location). This leads us to consider the following question: how
can we identify the \emph{best} location of the fragment in the genome?

This is the \emph{best location problem}. It ammounts to finding the
optimal alignment between two sequences, and for this reason has received
substantial attention in bioinformatics. It is relevant for the present
discussion because the candidate solution of the \emph{true} location
problem is always the \emph{best} location. The next section will further
clarify the relationship between the two problems.

\subsection{Seeding heuristics}

There exists exact algorithms to solve the best location
problem\cite{pmid7265238,pmid5420325}, but they are too slow to process
the large amount of data generated by modern sequencers. Instead, one uses
heuristic methods, \textit{i.e.} algorithms that run fast, but may not
return the best location\cite{Waterman1984}. Heuristic mapping algorithms
thus have three possible outcomes: $i.$ the DNA fragment is mapped to the
best location, $ii.$ the DNA fragment is mapped to a suboptimal location,
$iii.$ the DNA fragment is not mapped at all.

The most popular heuristic to address the best location problem is to
first identify short regions of high similiarity between the read and the
genome, and then run an exact alignment algorithm around those regions.
The first step is usually called ``seeding'' and the regions of high
similarity are called ``seeds''. This is for instance the general strategy
of the popular alignment algorithm BLAST\cite{pmid2231712}.

If seeds can be discovered fast, we can quickly zoom in on a small set of
candidate locations, and use slower exact algorithms on a reduced data
set. The disadvantage is that the target location is not always in the
candidate set. In this case, it cannot be discovered and the outcome is
either a false positive or a false negative.

This imposes a trade-off between speed and accuracy between each candidate
location must be tested with a slow algorithm. If the seeding step is set
to produce many candidates, the target is likely to be among them and it
will be discovered. The downside is that this will take long. Conversely,
if the seeding step is set to produce few candidates, the process will
terminate faster but the target is less likely to sit among the candidates
so it is less likely to be discovered.

Importantly, the terms of this trade-off are not set in stone. Some
seeding methods are better than others\cite{pmid16533404,pmid20460430}, so
one can develop faster mapping algorithms at no cost for accuracy.

This framework was developed for the best location problem, but it also
holds for the true location problem. Actually, the process is exactly the
same but the outcomes $i.$ and $ii.$ are different. The first, called a
``true positive'' or an ``on-target hit'', is that the DNA fragment is
mapped to the \emph{true} location; the second, called a ``false
positive'' or an ``off-target hit'' is that the DNA fragment is mapped
somewhere else in the genome. A true positive thus requires that the
target (the true location) is in the candidate set, and that it is the
optimum.

\subsection{MEM seeds}

In general, seeds do not have to be exact matches\cite{pmid20460430}. For
instance, the algorithm PatternHunter\cite{pmid11934743} uses ``spaced
seeds'' that tolerate mismatches\footnote{The method is actually covered
by a patent (reference US20070088510A1).}. However, in this document, we
will always assume that a seed is an exact match with no gap. In other
words, a seed is a perfect match between a part of the read and a genomic
site.

The rest of the document will focus on a seeding heuristic known as ``MEM
seeding''. This approach gives good empirical results when mapping short
reads (it is for instance the strategy\footnote{To be exact, BWA-MEM uses
SMEM seeds (Super Maximal Exact Match seeds).} of the popular mapper
BWA-MEM\cite{li2013aligning}) but there is presently no theory to justify
this performance. A deeper understanding of MEM seeding would thus be
useful to calibrate the speed/accuracy trade-off of this heurisitic.

Before developing a formal theory, let us clarify the meaning of the MEM
acronym.

\begin{definition}
A Maximal Exact Match (MEM) seed is a subsequence of a read that is
present in the genome of interest and that cannot be extended --- either
because the read ends or because the extended subsequence is not in the
genome.
\end{definition}

This definition presents a computational challenge: to know if a given
subsequence is a MEM seed, we need to know if it exists somewhere in the
genome. This is a non trivial problem in itself, but fortunately there
exists practical methods to identify all the MEM seeds from a read, even
for very large genomes\cite{ferragina2000opportunistic,
ferragina2005indexing}. Such algorithms are crucial for the present
discussion, but describing them falls outside the scope of the document.
We refer the interested reader to the relevant
literature\cite{pmid24336412,pmid25399029,pmid23349213,pmid19389736,
li2013aligning}, and here on we will assume that MEM seeds are simply
available at all times.

Before going further, we need to move a practical consideration out of the
way. It stands to reason that very short matches between the read and the
genome are uninformative. For instance, $99.7\%$ of all the possible
12-mers are present in the reference human genome, so an exact match of
size 12 or lower says nothing about the location of the DNA fragment.
In what follows, we will assume that subsequences of the read can only
match the original DNA sequence and/or its duplicates, but not the rest of
the genome. This will simplify the discussion without loss of generality
because non homologous matches are short and they can be ignored.

Fig.~\ref{fig:MEM_example} shows a concrete example. The target (the true
location of the read) is first potential location. The locations below
correspond to duplicate sequences. The mismatches between the genome and
the read are highlighted in bold. The read has three errors indicated by
stars, so \texttt{T}, \texttt{T} and \texttt{C} should actually be
\texttt{A}, \texttt{A} and \texttt{G}.

\begin{figure}[h]
\centering
\includegraphics[scale=1]{MEM_example.pdf}
\caption{\textbf{A concrete example of MEM seeding.}
A DNA fragment is sequenced with an imperfect instrument. It comes from
the top genomic location (Truth), but three errors were introduced
(indicated by stars). The sequence of the DNA fragment has three
duplicates at other genomic locations, shown below the target. Mismatches
between the read and the candidate sequence are highlighted in bold. MEM
seeds are indicated by grey boxes.}
\label{fig:MEM_example}
\end{figure}

The MEM seeds are indicated by grey boxes. Technically, they are part of
the read, but they are shown on the genome to highlight the position of
the match. One MEM seed cannot cover a smaller one (otherwise the smaller
would not be a MEM seed), but two consecutive MEM seeds can overlap, as in
the case of the frst two from the left. Note that overlapping MEM seeds
must match distinct sequences of the genome (otherwise neither of them
would be a MEM seed). Consecutive MEM seeds can also be disjoint, as in
the case of the second and the third from the left. In this case, they
must be separated by a nucleotide that is a mismatch against \emph{all}
the duplicates.

Observe that the rightmost MEM seed matches two genomic locations. This
case motivates the following definition, which will play an important role
for the development of the theory.

\begin{definition}
A \emph{strict} MEM seed has a single genomic match.
A \emph{shared} MEM seed has several genomic matches.
\end{definition}

Without surprise, the principle of MEM seeding is to use the genomic
matches of MEM seeds as candidate locations. Importantly, MEM seeds
shorter than a certain threshold $\gamma$ are ignored (in line with the
observation that short matches are uninformative). The parameter $\gamma$
is the tuning knob on the speed/accuracy trade-off. In
Fig.~\ref{fig:MEM_example} for instance, if $\gamma = 7$, there are three
candidate locations (the first, the second and the fourth) and if $\gamma
= 8$ there are only two (the first and the fourth).

Not every MEM seed makes the target a candidate location, so we introduce
the following terms to simplify the discussion.

\begin{definition}
Assume that the thresohld $\gamma$ is fixed. A \emph{good} seed is an
on-target MEM seed of size $\gamma$ or greater. Every other MEM seed is a
\emph{bad} seed. In other words, the true location is in the candidate set
if and only if the read has a good seed.
\end{definition}

Compared to seeds of fixed size, MEM seeds have at least two
counter-intuitive properties. The first is that there are cases where
there is no good seed, even when $\gamma$ is decreased.
Fig.~\ref{fig:full_masking_example} shows such an example. Even though
there is a single sequencing error, the read does not have any MEM seed
matching the target. Lowering $\gamma$ does not change this, so there is
no way to discover the true location (even though it is also the best
location).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{full_masking_example.pdf}
\caption{\textbf{MEM masking case 1.}
The read, the MEM seeds and the sequences are as in
Fig.~\ref{fig:MEM_example}. The MEM seeds matching the two incorrect
locations at the bottom effectively mask the target, so it cannot
be discovered. MEM masking can occur even when the true location is the
best and when there is a single error on the read.}
\label{fig:full_masking_example}
\end{figure}

The second counter-intuitive property of MEM seeding is that there are
cases where there would be a good seed if the read were shorter.
Fig.~\ref{fig:short_vs_long} shows an example of this case. Again, there
is no on-target MEM seed, but there would be if the read were two
nucleotides shorter on the right side. Indeed, in this case there would be
a shared MEM seed matching the first and the second locations. It would
also be a good seed for $\gamma \leq 12$.

\begin{figure}[h]
\centering
\includegraphics[scale=1]{short_vs_long_example.pdf}
\caption{\textbf{MEM masking case 2.}
The read, the MEM seeds and the sequences are as in
Fig.~\ref{fig:MEM_example}. There is no on-target MEM seed, but there
would be a shared MEM seed if the read were two nucleotides shorter. The
true location has been masked by sequencing the last two nucleotides.}
\label{fig:short_vs_long}
\end{figure}

These examples show that MEM seeds can perform worse than seeds of fixed
size, again raising the question why they seem to give good performance in
practice. The theory developed below will allow us to compute the
probability that a read has no good seed and thus that the true location
is missed. This probability is the key to evaluating and calibrating the
MEM seeding heuristic.

Before going further, we recommend the readers that are not already
familiar with MEM seeding to take a moment to study the examples above.
The complexity of the next sections will go \textit{crescendo} and they
will draw heavily from the concepts and definitions above.

@@

In this section we represent reads as combinatorial ojects. The
construction has two essential purposes: the first is to give a generative
model of reads that do not have any on-target MEM seed; the second is to
allow us to find the weighted generating function of those reads, which in
turn will allow us to find their probability of occurrence.

@@

The nucleotide sequence of the read is inadequate to know where MEM seeds
are and whether they are on target. For this purpose, we will recode reads
as sequences of letters from the four letter \emph{MEM alphabet}
$\{\downarrow, \uparrow, |, \square\}$.

Each nucleotide is recoded with $\downarrow$, $\uparrow$, or $\square$.
The $\downarrow$ symbol indicates that the nucleotide is a sequencing
error. By definition, it is a mismatch for the target. The $\uparrow$
symbol indicates that the nucleotide is the ``first'' nucleotide of an
on-target MEM seed, which we are about to clarify. All the other
nucleotides are represented by the $\square$ symbol. Finally, the $|$
symbol is appended after the last nucleotide of the read. It is an
obligatory terminator that does not correspond to an actual nucleotide.

To explain the precise meaning of the $\uparrow$ symbol, let us consider
the left end of the read: as we try to extend the first match as much as
possible to the right, the number of hits in the genome decreases. It may
be the case that at some point, there is a single hit and that it is on
target. The very first nucleotid where that happens is called the
\emph{pivot} and is labelled with the symbol $\uparrow$. An on-target MEM
seed has exactly one pivot, so the number of $\uparrow$ symbols is the
number of on-target MEM seeds in the read. Note that $\uparrow$ symbols
marks all on-target MEM seeds, regardless of their size.

Fig.~\ref{fig:sketch_MEM} shows the read of Fig.~\ref{fig:MEM_example} as
encoded with the MEM alphabet. Below the read are the matches between the
read and the potential locations from Fig.~\ref{fig:MEM_example}. A black
square means that the nucleotide is a mismatch for the genomic sequence,
and a white or grey square means that it is a match. Grey squares are used
to highlight MEM seeds, as in Fig.~\ref{fig:MEM_example}.

\begin{figure}[h]
\centering
\includegraphics[scale=.85]{sketch_MEM.pdf}
\caption{\textbf{The MEM encoding}.
The read of Fig.~\ref{fig:MEM_example} is recoded with the MEM alphabet.
The mismatches between the read and the potential locations are indicated
with black boxes and the matches with white boxes. The grey boxes are
matches and they indicate the locations of the MEM seeds. The $\downarrow$
symbols coincide exactly with the mismatches between the read and the
target. The $\uparrow$ symbol is always a match for the target and a
mismatch for some off-target location. It indicates the presence of a
strict on-target MEM seed (flanked by the nearest $\downarrow$ symbols or
by the end of the read). Also note the presence of the $|$ terminator at
the end of the read.}
\label{fig:sketch_MEM}
\end{figure}

Observe that the two on-target MEM seeds contain a pivot, but the shared
MEM seed does not. This is always the case. Recall that the pivot
indicates that the true location is the \emph{only} local match, so the
$\uparrow$ symbol cannot appear in on-target MEM seeds when they are
shared.

The MEM alphabet makes it easy to find strict on-target MEM seeds: they
are the longest stretches of symbols containing $\uparrow$ and not
containing $\downarrow$. However, shared on-target MEM seeds do not appear
in this representation.

@@

To make shared on-target MEM seeds apparent, we decorate the $\uparrow$
and $\downarrow$ symbols, effectively extending the alphabet. The
$\downarrow$ symbol is replaced by the set of symbols $\downarrow_{/m}$,
where $m$ is the number of duplicates that \emph{match} the erroneous
nucleotide. The $\uparrow$ symbol is replaced by the set of symbols
$\uparrow_{/j}$, where $j-1$ is the number of consecutive $\square$
symbols preceding $\uparrow$.

Note that the numbers decorating the symbols have a very different nature.
In the case of the $\downarrow$ symbol, it is an information about the
number of sequences matching the read; in the case of the $\uparrow$
symbol, it is the position of the pivot relative to the previous
sequencing error (or relative to the beginning of the read).

\begin{figure}[h]
\centering
\includegraphics[scale=.85]{sketch_extended.pdf}
\caption{\textbf{The extended MEM encoding}.
The read of Fig.~\ref{fig:sketch_MEM} is represented in the extended MEM
alphabet. The symbols $\uparrow$ and $\downarrow$ are now decorated by
numbers. The arrows departing from those numbers help understand their
meaning. The $\downarrow$ symbol is decorated by the number of locations
that match the nucleotide. The $\uparrow$ symbol is decorated by the
number of consecutive $\square$ symbols before it, plus one.}
\label{fig:sketch_extended}
\end{figure}

Fig.~\ref{fig:sketch_extended} shows the encoding of the read from
Fig.~\ref{fig:sketch_MEM} in the extended MEM alphabet. The arrows bring
the focus on how the numbers decorating the $\uparrow$ and $\downarrow$
symbols are computed. For instance, the $\downarrow_{/0}$ symbol indicates
that the nucleotide is an error (this is the meaning of the $\downarrow$
symbol), and that it matches no other location (this is meaning of $/0)$.
In other words, it is a mismatch against every potential location.

In this new encoding, shared on-target MEM seeds are stretches of
$\square$ symbols flanked either by $\downarrow_{/0}$, or by $|$, or by
the left end of the read. Indeed, such a stretch is a MEM seed because it
matches at least one sequence (the target) and it cannot be extended
(because the end of the read is reached or because there is no match in
the genome). Also, it cannot be a strict on-target MEM seed because it
does not contain the $\uparrow$ symbol, so it must be a shared on-target
MEM seed.

The reason for introducing more than just the $\downarrow_{/0}$ symbol is
not immediately clear. For the combinatorial construction to be useful, we
need not only to identify on-target MEM seeds, but also to compute the
probabilities of the reads of each type. This is precisely the role of the
extended alphabet, as we will see in section~\ref{sec:transfer_mat}.

\subsection{Reads as extended MEM segments}

The final twist in the combinatorial construction is to consider reads not
as sequences of symbols, but as sequences of \emph{segments}. The purpose
here is merely to express that some on-target MEM seeds are good
(\textit{i.e.}, their size is $\gamma$ or greater) and some are not.

\begin{definition}
\label{def:segment}
A segment is a maximal sequence of $\square$ symbols flanked on the right
side by one of the symbols $\uparrow_{/j}$, $\downarrow_{/m}$, or $|$. The
one and only segment flanked by $|$ is called the tail. It is the only
segment that can have size 0.
\end{definition}

In line with definition~\ref{def:segment}, the symbols $\uparrow_{/j}$,
$\downarrow_{/m}$ and $|$ are referrred to as \emph{terminators}.
Fig.~\ref{fig:sketch_segment} shows the segment representation of the read
from Fig.~\ref{fig:sketch_extended}. The numbers below indicate the size
of each segment.

\begin{figure}[h]
\centering
\includegraphics[scale=.85]{sketch_segments.pdf}
\caption{\textbf{The extended MEM segment encoding}.
The read from Fig.~\ref{fig:sketch_segment} is represented as a sequence
of segments. Each segment is terminated by either $\uparrow_{/j}$, or
$\downarrow_{/m}$ or $|$. The rightmost fragment is the tail. The numbers
below indicate the size of each segment in nucleotides.}
\label{fig:sketch_segment}
\end{figure}

Shared on-target MEM seeds at the tail of the read are segments but all
other on-target MEM seeds are not. Shared MEM seeds cannot contain the
terminator $\downarrow_{/0}$ because it is a mismatch against the target.
Strict on-target MEM seeds are split over at least two segments: one
terminated by $\uparrow_{/j}$ and one right of it terminated by either
$\downarrow_{/m}$ or $|$.

@@

We now turn to the question of translating the extended MEM segment
representation to a construction of reads without good seeds
(\textit{i.e.}, without on-target MEM seeds of size $\gamma$ or greater).
For this, we use tools borrowed from analytic combinatorics. We give the
minimum amount of background that is necessary to expose the theory. A
more detailed account of the strategy to compute seeding probabilities
with analytic combinatorics has been published
elsewhere\cite{filion2017analytic,filion2018analytic}. We refer the
readers interested in learning more about analytic combinatorics to the
textbooks\cite{flajolet2009analytic,sedgewick2013introduction}.

@@

Let $\mathcal{A}$ be a set of combinatorial objects with a size $k \in
\mathbb{N}$ and a probability of occurrence. Denoting $a_k$ the total
probability of objects of size $k$, the function defined as $A(z) =
\sum_{k\geq0} a_kz^k$ is called the \emph{weighted generating
function} of the combinatorial objects.

In our case, we are interested in the weighted generating function of
reads without good seeds, so the coefficient $a_k$ is the probability
that a read of size $k$ has no good seed. As $k$ becomes larger,
it becomes more difficult to evaluate $a_k$, but the representation
developed above will allow us to do this.

@@

The combinatorial construction of reads using the extended MEM alphabet
gives us a method to express their weighted generating function from the
simpler generating function of segments. More specifically, we can
describe complex sequences of segments through a so-called \emph{transfer
matrix} $M(z)$ that specifies which pairs of consecutive segments are
allowed and which are forbidden in the reads of interest.

The segment types are identified by their terminators arranged in a
predefined oder. The entry of $M(z)$ at coordinates $(i,j)$ is the
weighted generating function of segments with the $j$-th terminator that
can be inserted immediately after segments with the $i$-th terminator. The
entries of the matrix $M(z) + M(z)^2 + \ldots = M(z) \cdot (I-M(z))^{-1}$
are the weighted generating functions of the reads of all sizes using all
the allowed combinations of segments.

The entry of interest from $M(z)$ is the one that corresponds to the
terminators $\downarrow_{/0}$ and $|$ because it represents the reads of
any size whose first segment can be inserted after $\downarrow_{/0}$, and
whose last segment is terminated by $|$. Indeed, the $|$ terminator marks
the end of the read, so it must be the final symbol of every valid read.
As for the first segment, observe that all the segments that can follow
the $\downarrow_{/0}$ symbol are exactly those that can be at the frst
postion of the read\footnote{One could have introduced a special start
symbol. It is equivalent to an invisible $\downarrow_{/0}$ prepending
the reads.}.

In summary, the transfer matrix $M(z)$ contains the weighted generating
function of interest. Our next task is thus to specify $M(z)$.

\section{The transfer matrix}
\label{sec:transfer_mat}

\subsection{Hard and soft masking}
